Synthesizing Verified Components for Cyber Assured Systems Engineering \thanks{DISTRIBUTION STATEMENT A. Approved for public release.}

### 2022-03-28 paper/agree-semantics.tex
Condensing Sec 4.2; added table mapping math defs to AGREE. Some entries look redundant

The semantics of terms and formulas, as mentioned in the previous
section, is in terms of \emph{streams of values}. Values encompass at
least booleans and numbers, but can be readily extended to include
records and arrays. A value stream is a total function from time
(natural numbers including 0) to values:
\[
 \konst{stream} = \mathbb{N}_0 \to \konst{value}
\]
Given an \emph{environment} $E : \konst{name} \mapsto \konst{stream}$
binding variable names to value streams, the semantics $\sem{-}^E_t$
of terms and formulas defines the meaning of compound syntax in terms
of the meaning of subexpressions. The value of a variable $v$ at time
$t$ is found by looking up the stream bound to $v$ in $E$ (call it
$s$) and returning $s_t$. Some clauses of the semantics follow,
omitting the temporal operators:
\[
\begin{array}{rcl}
\sem{v}^E_t & = & E(v)(t) \\
\sem{c}^E_t & = & c \\
\sem{e_1 + e_2}^E_t & = & \sem{e_1}^E_t + \sem{e_2}^E_t \\
   & \cdots & \\
\sem{b_1 \land b_2}^E_t & = & \sem{b_1}^E_t \land \sem{b_2}^E_t \\
   & \cdots & \\
\end{array}
\]

The temporal operators deal with time in more significant ways. The
value of $\konst{pre}(e)$ at time $t$ is the value of $e$ at time
$t-1$ (at time zero, \konst{pre} is undefined).
A followed-by, $e_1 \to e_2$, evaluates to $e_1$ at time zero; otherwise it is $e_2$.
Followed-by is often used to give a value to a stream in the first step so that subsequent steps are able to use $\konst{pre}(e)$.
\[
\begin{array}{rcl}
\sem{\konst{pre}(e)}^E_t & = & \sem{e}^E_{t-1}, \mathrm{when}\ t > 0 \\
\sem{e_1 \to e_2}^E_t & = & \itelse{t=0}{\sem{e_1}^E_0}{\sem{e_2}^E_t} \\
\sem{\konst{Hist}(b)}^E_t & = & \forall n \leq t.\; \sem{b}^E_n
\end{array}
\]

@@

>>>>

\subsection{Semantics}
The semantics of terms and formulas, as mentioned in the previous
section, is in terms of \emph{streams of values}. Values encompass at
least booleans and numbers, but can be readily extended to include
records and arrays. A value stream is a total function from time
(natural numbers including 0) to values:
\[
 \konst{stream} = \mathbb{N}_0 \to \konst{value}
\]
Given an \emph{environment} $E : \konst{name} \mapsto \konst{stream}$
binding variable names to value streams, the semantics $\sem{-}^E_t$
of terms and formulas defines the meaning of compound syntax in terms
of the meaning of subexpressions. The value of a variable $v$ at time
$t$ is found by looking up the stream bound to $v$ in $E$ (call it
$s$) and returning $s_t$. Some clauses of the semantics follow,
omitting the temporal operators:
\[
\begin{array}{rcl}
\sem{v}^E_t & = & E(v)(t) \\
\sem{c}^E_t & = & c \\
\sem{e_1 + e_2}^E_t & = & \sem{e_1}^E_t + \sem{e_2}^E_t \\
   & \cdots & \\
\sem{b_1 \land b_2}^E_t & = & \sem{b_1}^E_t \land \sem{b_2}^E_t \\
   & \cdots & \\
\end{array}
\]

The temporal operators deal with time in more significant ways. The
value of $\konst{pre}(e)$ at time $t$ is the value of $e$ at time
$t-1$ (at time zero, \konst{pre} is undefined).  A followed-by, $e_1
\to e_2$, evaluates to $e_1$ at time zero; otherwise it is $e_2$.
Followed-by is typically used to give a value to a stream in the first
step so that subsequent steps are able to use $\konst{pre}(e)$.
\[
\begin{array}{rcl}
\sem{\konst{pre}(e)}^E_t & = & \sem{e}^E_{t-1}, \mathrm{when}\ t > 0 \\
\sem{e_1 \to e_2}^E_t & = & \itelse{t=0}{\sem{e_1}^E_0}{\sem{e_2}^E_t} \\
\sem{\konst{Hist}(b)}^E_t & = & \forall n \leq t.\; \sem{b}^E_n
\end{array}
\]

@@

\paragraph{AGREE notation.}
AGREE uses ASCII notation for the mathematical syntax we have
defined. The following table spells out the mapping.
\[
\begin{tabular}{l|l|l}
 Operation & Math. notation & AGREE \\ \hline
 followed-by &  $e \to e$          & e \verb+->+ e  \\
 logical implication &  $e \imp e$ & e \verb+=>+ e \\
 if-and-only-if &  $e \iff e$      & e \verb+<=>+ e \\
 variable definition & $v = e$     & \konst{eq} v\ e \\
 Historically  & $\konst{Hist}(b)$ & \konst{Historically}(b) \\
\end{tabular}
\]

### 2022-03-28 paper/sw-contract.tex
Condensing Sec 4.2; added table mapping math defs to AGREE. Some entries look redundant

The actual surface syntax for AGREE that is embedded in OSATE is slightly different than the formal semantics in the previous section.
The mapping between the two is straightforward.
The AGREE specification for the SW component in the example of
Section~\ref{sec:example} is given in \figref{fig:sw}.  The
specification uses \texttt{eq} statements to define variables local to
the contract specification.  For
example, \lineref{line:sw-event-def-start} defines the \texttt{req}
variable to be equivalent to the \texttt{event} expression.

All the named ports in the corresponding AADL component are in the scope of the specification, and there are additional implicit boolean \emph{event} inputs (or outputs) associated with event ports.
An \texttt{event} expression refers to that implicit input (or output) boolean value and is true when data is present on the named port and false otherwise.
\linesref{line:sw-event-def-start}{line:sw-event-def-end} create local variables that are true when data is present on the corresponding event ports for the component.
The local variables here are purely for convenience in writing the specification.

The \texttt{assume} statement is a string description followed by a predicate.
Those on \lineref{line:sw-assume-1} and \lineref{line:sw-assume-2} are implications requiring that when data is present it is well-formed.
The well-formed predicates themselves are defined elsewhere using AGREE functions.

The assumption on \lineref{line:sw-assume-3} constrains when a request
can arrive by reasoning about the \emph{state} of the contract defined
by the associated streams.  When writing assumptions and guarantees,
it is important to differentiate pre-state, before a contract updates
its state, and post-state, after a contract updates its state, in
response to the current input.  The \texttt{pre} naturally makes that
distinction.  In general, assumptions regarding state should be
evaluated in the pre-state of the component with the current inputs,
and guarantees regarding state should be evaluated on the post-state
of the component given the current inputs.  Additionally, guarantees
should use \texttt{pre} anytime they need to reason about the current
state relative to the previous state.

The assumption on \lineref{line:sw-assume-3} uses a followed-by expression so it is \texttt{true} at time zero and then it is the truth value of the implication in all future instances.
The followed-by guards the \texttt{pre} as mentioned previously;
thus after time zero, the assumption depends on the presence of a request and the pre-state of the component.
The assumption on \lineref{line:sw-assume-3} is that if there is an incoming request, it is either the very first one, \texttt{\textbf{Historically}(\textbf{not} req)}, or it is after the component has output a start event in response to a previous request, \texttt{\textbf{Since}(\textbf{not} req, strt)}---\emph{not request since start}.

The guarantees on \linesref{line:sw-guarantee-1}{line:sw-guarantee-3} coincide events and assert well-formed output.
The guarantees on \linesref{line:sw-guarantee-4}{line:sw-guarantee-6} define temporal properties of the component.
\lineref{line:sw-guarantee-4} insists that a start happens with a request or one step after a request.
The guarantee uses the assumption on \lineref{line:sw-assume-3} and does not check for two requests in a row without a start as the assumption precludes that input behavior.
It differentiates with the followed-by what is required at time zero, the start must coincide with the request, with what is required after time zero, the start must coincide with the request or is a response to a request one step earlier.
The guarantee also does not force the start to always happen, it only says that if it does happen, it is under the defined conditions.

\lineref{line:sw-guarantee-5} forces the alert to sound if the start does not arrive within the one-step bound.
Together with \lineref{line:sw-guarantee-4} the contract model allows for non-alerting and alerting behavior.
\lineref{line:sw-guarantee-6} ensures if alert has ever happened, then it is happening in the present moment.
These six guarantees define the behavior of the SW component under the three assumptions and correspond to the informal descriptions given in \secref{sec:example}.

>>>>

%% The actual surface syntax for AGREE that is embedded in OSATE is
%% slightly different than the formal semantics in the previous section.
%% The mapping between the two is straightforward.

%% The
%% specification uses \texttt{eq} statements to define variables local to
%% the contract specification.  For
%% example, \lineref{line:sw-event-def-start} defines the \texttt{req}
%% variable to be equivalent to the \texttt{event} expression.

%% All the named
%% ports in the corresponding AADL component are in the scope of the
%% specification, and there are additional implicit boolean \emph{event}
%% inputs (or outputs) associated with event ports.  An \texttt{event}
%% expression refers to that implicit input (or output) boolean value and
%% is true when data is present on the named port and false otherwise.

%% An \texttt{assume} statement is a string description followed by a
%% predicate.

%% When writing assumptions and guarantees,
%% it is important to differentiate pre-state, before a contract updates
%% its state, and post-state, after a contract updates its state, in
%% response to the current input.  The \texttt{pre} naturally makes that
%% distinction.  In general, assumptions regarding state should be
%% evaluated in the pre-state of the component with the current inputs,
%% and guarantees regarding state should be evaluated on the post-state
%% of the component given the current inputs.  Additionally, guarantees
%% should use \texttt{pre} anytime they need to reason about the current
%% state relative to the previous state.

%% This condition constrains when a request
%% can arrive by reasoning about the \emph{state} of the contract defined
%% by the associated streams. The assumption uses a followed-by
%% expression so it is \texttt{true} at time zero and then it is the
%% truth value of the implication in all future instances.  The
%% followed-by guards the \texttt{pre} as mentioned previously; thus
%% after time zero, the assumption depends on the presence of a request
%% and the pre-state of the component.  The assumption
%% on \lineref{line:sw-assume-3} is that

The AGREE specification for the SW component in
Section~\ref{sec:example} is given in \figref{fig:sw}.
\linesref{line:sw-event-def-start}{line:sw-event-def-end} define
local variables that are true when data is present on the
corresponding event ports for the component.  These variables are
purely for notational convenience. The contract has three assumptions:
\begin{itemize}
\item \lineref{line:sw-assume-1} and \lineref{line:sw-assume-2}.
When input is present it is well-formed. The well-formed predicates
are defined elsewhere using AGREE functions.

\item \lineref{line:sw-assume-3}. If there is an incoming request,
it is either the very first one, \texttt{\textbf{Historically}(\textbf{not} req)},
or it arrives after the component has output a start event in response to a previous
request, \texttt{\textbf{Since}(\textbf{not} req, strt)}---to be read as \emph{not
request since start}.
\end{itemize}

The contract has six guarantees:
\begin{itemize}
\item \linesref{line:sw-guarantee-1}{line:sw-guarantee-3}
coincide events and assert well-formed output.\footnote{\konst{KLS}:I am confused about the ``coincide events''.}
\item \lineref{line:sw-guarantee-4} insists that a start happens with a request or one step after a request. At time zero the start must coincide
with the request; after that, the start must coincide with the request
or else is a response to a request one step earlier.  The guarantee
works with the assumption on \lineref{line:sw-assume-3} and thus does
not have to check for two requests in a row without a start, since the
assumption precludes that input behavior. The guarantee also does not
force the start to always happen, it only says that if it does happen,
it is under the defined conditions.

\item \lineref{line:sw-guarantee-5} forces the alert to sound if the start does not arrive within the one-step bound. Together with \lineref{line:sw-guarantee-4} this allows
for non-alerting and alerting behavior.
\item \lineref{line:sw-guarantee-6} guarantees that, if an alert has ever happened,
it continues to happen.
\end{itemize}
These six guarantees define the behavior of the SW component under the
three assumptions and correspond to the informal descriptions given
in \secref{sec:example}.
